Machine Learning for Mechanistic Models of Metapopulation Dynamics

### 2023-11-10 ms.Rnw
update profile, filter_tests, and some edits

<<setup,echo=F,message=F>>=
# computational intensity, i, also known as run level
i <- 3
library(metapoppkg)
Ncities <- ncol(metapoppkg::incidence) - 1 # first column is date
U <- switch(i,20,Ncities,Ncities)
library(foreach)
library(doParallel)
library(doRNG)
cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
if(is.na(cores)) cores <- detectCores()
registerDoParallel(cores)

@@

\begin{abstract}
Mathematical models in ecology and epidemiology demand data fidelity for reliable knowledge and sound policy. Metapopulation systems, involving sub-populations across various locations, pose technical challenges in statistical inference due to nonlinear, stochastic interactions. These challenges can divert attention from core scientific inquiries concerning the link between mathematical models and scientific understanding. Progress in statistically efficient simulation-based inference for partially observed stochastic dynamic systems has enabled the development of statistically rigorous approaches to the analysis of nonlinear but low-dimensional systems. Recently, an algorithm has been developed which enables comparable inference for higher-dimensional models arising in metapopulation systems. The COVID-19 pandemic provides a situation where mathematical models and their policy implications were widely visible, and we revisit an influential metapopulation model and dataset used to inform basic epidemiological understanding early in the pandemic. Our methods support self-critical data analysis, enabling us to identify and address weaknesses, and leading to a new model with substantially improved statistical fit and parameter identifiability. Our findings reveal that the January 23, 2020 lockdown in China was more effective than previously thought. We also discuss recommended statistical analysis standards for future metapopulation system modeling.
\end{abstract}

@@

Biological populations may be structured into a collection of densely-populated communities separated by sparsely populated regions.
The communities, which may be cities in a human context, comprise a metapopulation.
Motivation for metapopulation modeling arises when some essential feature of the population dynamics cannot be understood from looking at a single location.
Dynamics of persistence through local extinctions and reintroductions have been extensively studied in ecology \cite{hanski98,mackenzie09}.
In epidemiology, metapopulation dynamics can be a barrier to the eradication of a pathogen, and may determine the successful invasion of a new pathogen or a new strain of an existing pathogen \cite{metcalf21}.
In other situations, spatiotemporal dynamics may be an unavoidable component of the system under study without being the focus of the investigation \cite{zhang22,wheeler23}.

A recent growth of research in metapopulation dynamics, developing mathematical models and fitting them to spatiotemporal data, has been driven partly by the COVID-19 pandemic \cite{li20,wu20,wang22,prieto22,cascante-vega22,pizzuti20,alleman21,yang21,engebretsen23} and in part by methodological advances.
Developing dynamic models with both scientific and statistical justification was a longstanding open problem for even a single community, until the start of this millenium \cite{bjornstad01}.
Over the past two decades, new algorithms \cite{ionides06,toni09,andrieu10,ionides15} and software \cite{king16,kristensen16,devalpine17} together with ever-increasing computational resources, have enabled routine inference for low-dimensional nonlinear partially observed stochastic dynamic systems.
However, fundamental algorithmic scalability issues known as the ``curse of dimensionality'' lead to difficulties with the high-dimensional systems arising in metapopulation inference.
These issues are clearest for Monte Carlo techniques based on importance sampling \cite{bengtsson08} but are also evident in the need for variational approximations for large Monte Carlo Markov Chain (MCMC) calculations \cite{blei17}. 
The quest for numerical approximations suitable for metapopulation models of scientific interest has excited the development of various methodological approaches.
Nevertheless, limitations have persisted in the capability to carry out flexible model-based inference and rigorous model criticism.
Thus, data analysis for metapopulation models has lagged behind low-dimensional time series analysis of biological systems.
Recent developments enable this gap to be closed, as we demonstrate via a reanalysis of COVID-19, viewed from the context of the ability to drawm evidence-based scientific conclusions about the dynamics of the emerging pandemic in January and February 2020.

@@

A natural place to look for statistical methodology applicable to metapopulation models is among the techniques developed for inferring population dynamics at a single location, reviewed by \cite{funk20,auger-methe21}.
Commonly implemented approaches for POMP models can be categorized as  (i) MCMC; (ii) matching summary statistics between data and simulations; (iii)  linearization;   (iv) particle filters (sequential Monte Carlo). For low-dimensional systems, MCMC is generally possible and have the attractive feature of statistically efficient use of data, i.e., accessing the likelihood function for Bayesian or likelihood-based inference, with approximation resulting only from finite Monte Carlo effort.
In principle, MCMC techniques enable Bayesian inference or maximum likelihood via expectation-maximization algorithms \cite{cappe05}.
For high-dimensional systems, scalability considerations demand further approximations. 
In practice, successful MCMC for metapopulation models requires careful model-specific algorithm development \cite{whitehouse23}.
Another approach to fitting arbitrary metapopulation models involved matching summary statistics of simulations to the corresponding data statistic.
In the context of Bayesian inference this is called Approximate Bayesian Computing \cite{conlan12}.
However, informative, low-dimensional summary statistics can be hard to construct even for low-dimensional nonlinear systems \cite{fasiolo16,shrestha11}.

@@

Particle filters, which are applicable to a flexible class of models for low-dimensional nonlinear systems \cite{breto09,king16} suffer acutely from the ``curse of dimensionality'' \cite{bengtsson08}.
In contrast, block particle filter (BPF) methods achieve scalability by updating particles through localized resampling, rather than employing the linear update of EnKF \cite{rebeschini15}.
It is an empirical question which of these approximations is more successful on metapopulation models,  with prior evidence favoring BPF \cite{ionides21}.
In this paper, we not only aim to compare these methods in the context of metapopulation modeling but importantly also we reveal issues of model assessment when fitting complicated models to large datasets.
We introduce a model diagnosis procedure to mitigate the risk of selecting an inappropriate model, an essential step in informing public policy decisions.

@@

As an example, we reconsider the influential analysis of COVID-19 from early in the pandemic by Li et al. \cite{li20}, that provided estimates of transmission parameters and the effect of lockdown using the limited data available at the time.
Other teams have fitted models to address similar questions \cite{kraemer20,yang21,brett23} but the study by \cite{li20} is distinctive for fitting a stochastic mechanistic metapopulation model to extensive spatiotemporal data for 373 Chinese cities.
The results were published in May, 2020, based on reported cases from January 10 to February 8 of that year.
The state-of-the-art spatiotemporal analysis was possible on an urgent timescale because the team of researchers had developed their methodology in a sequence of previous situations \cite{shaman12,yang14,yang15,pei18}.
The paper is written with attention to reproducibility, and the main results are strengthened by various supporting analyses in an extensive supplement.
While examining the points mentioned above, we have identified various weaknesses that could have been mitigated by adhering to the aforementioned recommendations.
Our goal is not to criticize any specific paper, but rather to build on the timely analysis of \cite{li20} to demonstrate how recently developed techniques provide possibilities to carry out superior data analysis in future.

@@

\end{tikzpicture}
}
\end{center}
\vspace{-5mm}
\caption{A flow diagram for the SEAIR metapopulation model. Each individual in city $u$ is a member of exactly one of the square blue compartments. Individuals entering the reportable infectious compartment, $\mathrm{I}_u$ for city $u$, are simultaneously included in the delayed reporting process compartment, $\mathrm{C^a_u}$. Upon arrival at the final reporting compartment, $\mathrm{C}_u$, the individual is included in the case report for city $u$. Movement of individuals between cities occurs by transport to and from a transport compartment, $\mathrm{T}$. Movement is modeled only for susceptible, exposed, and undetected infections.}\label{fig:flow_diagram}
\end{figure}

We consider metapopulation models where the units are 373 provincial cities in China (meaning cities with administrative responsibility for an entire region) and the data are daily reported COVID-19 cases. 
COVID-19 dynamics are represented by an Susceptible-Exposed-Asymptomatic-Infectious-Recovered (SEAIR) epidemic spreading model. 
Questions of urgent interest early in the pandemic include the relative transmissibility of reported to unreported cases, the fraction of unreported cases, and the effect on transmission of movement restrictions imposed on and around January 23 \cite{li20}.
The model structure is illustrated by the flow diagram in Fig.~\ref{fig:flow_diagram}.
We consider different model implementations within this structure, with full model specifications provided by the equations and parameter values in Supplementary Sec.~\suppSecOurModel.

Our starting point is model {\LiMobility} which is based on the model of \cite{li20}, with specific differences discussed in Supplementary Sec.~\suppSecLiModel.
We consider the full dataset, from January 10 to February 8, with transmission parameters re-estimated following the lockdown on January 23; these correspond to the periods~1 (January 10 to January 22) and 3 (January 24 to February 8) of \cite{li20}.
Simulations from {\LiMobility}, using the parameters of \citep{li20}, closely match simulations from the code provided by \cite{li20} (Supplementary Sec.~\suppSecLiModel).
The minor differences between the results of {\LiMobility} and \citep{li20} enable us to place their model within the general metapopulation framework described by \citep{ionides21}.
Inspection of the mobility data reveals that some small cities have no recorded incoming travelers, and therefore no possibility within the model of a SARS-CoV2 introduction (Supplementary Sec.~\suppSecMobility).
This minor limitation formally results in a likelihood of zero for the simulation model (i.e., it is impossible for the simulation model to reproduce the observed spatiotemporal dataset), and hence a log-likelihood of $-\infty$.

@@

\begin{table}[ht]
\small
\begin{tabular}{c|rr|l}
Model & loglik & df & description
\\
\hline
{\LiMobility} &
  $-\infty$ & 10 &
  SEAIR model using the parameter values and mobility data of \cite{li20}
\\
{\LiParams} &
  \Sexpr{myround(mean(M2_loglik['est']),1)}\stdError{\Sexpr{myround(M2_loglik['se'],1)}}& 10 &
  Adjusted mobility and measurement in {\LiMobility}
\\
{\BenchmarkIID} &
  \Sexpr{myround(-iid_mle3$value,1)}  & \Sexpr{U+1} &
  Independent identically distributed negative binomial
\\
{\BenchmarkAR} &
 \Sexpr{myround(-ar_mle3$value,1)}  & \Sexpr{U+2} &
  Autoregressive negative binomial
\\
{\RevisedModelUnconstrained} &
  \Sexpr{myround(lik3["logLik"],1)}\stdError{\Sexpr{myround(lik3["logLik_se"],1)}}&
  12 &
  Adding overdispersed dynamics to {\LiParams} and refitting
\\
{\RevisedModelConstrained} &
  \Sexpr{myround(lik6["logLik"],1)}\stdError{\Sexpr{myround(lik6["logLik_se"],1)}}&
  10 &
  Latent and infectious durations unchanged by lockdown in {\RevisedModelUnconstrained}.
\\

@@

We avoided the issue of model {\LiMobility} by adding some additional mobility based on a gravity movement model, as described in Supplementary Sec.~\suppSecMobility, giving rise to model~{\LiParams}.
We implemented an additional adjustment between models {\LiMobility} and {\LiParams} to align the measurement model with the ensemble Kalman filter (EnKF) inference method presented by \cite{li20}.
That EnKF implementation involved specifying a quanitity called the observation error variance, defined as a function of the observed cases, to quantify the uncertainty in the measurements.
Within the POMP specification, the measurement variance can depend on the latent state but not directly on the observed data.
To interpret the choice of EnKF observation variance within the POMP framework, we specified the measurement model for~{\LiParams} to have equivalent scaling to the choice of \citep{li20}, but with dependence on the reported cases replaced by dependence on the modeled, but unobserved, exact case count.

Based on a comparison of various nonlinear spatiotemporal filters (Supplemtary Fig.~\suppFigFilterComparison) we evaluated the log-likelihood for {\LiParams} using a block particle filter, reported in Table~\ref{tab:loglik}. To account for model overfitting, the number of estimated parameters can be subtracted from the log-likelihood to obtain a comparison equivalent to Akaike's Information Criterion (AIC) \citep{burnham02}.
When the difference in log-likelihood is large compared to the difference in degrees of freedom, the ordering of statistical goodness-of-fit is clear without presenting formal statistical hypothesis tests.

To find out whether this log-likelihood value suggests that the model is satisfactory, we compare it with two simple statistical models:
{\BenchmarkIID} simply models the daily case report for each city as an independent identically distributed (IID) negative binomial random variable; 
{\BenchmarkAR} adds an autoregressive component to {\BenchmarkIID} (see Supplemetary Sec.~\suppSecBenchmark).
We see from Table~\ref{tab:loglik} that both {\BenchmarkIID} and {\BenchmarkAR} outperform {\LiParams} by many units of log-likelihood.
Likelihood can properly be compared between different models for the same data, with statistical uncertainty in log-likelihood differences arising on the unit scale \citep{pawitan01}.
When the fit of a mechanistic model is inferior to a simple statistical model, we learn that the mechanistic model has room for improvement as a description of the data, but we do not immediately learn what the deficiency is.
The development of methods for formal statistical fitting of mechanistic models has led to increased understanding of the importance of modeling variability in the dynamics \cite{he10,stocks20,whitehouse23}.
We therefore hypothesized that the fit of {\LiParams} could be improved by permitting additional dynamic stochasticity.

@@

All parameter estimates are reported in Supplementary Table~\suppTableResults.
Here, we discuss the estimated basic reproductive number (i.e., the expected number of secondary infections from one index case in a fully susceptible population), denoted by $\Rzero^{\before}$ and $\Rzero^{\after}$ before and after the January 23rd lockdown.
Our estimates for model {\RevisedModelConstrained}, are $\Rzero^{\before}=\Sexpr{myround(ci_c["R0_be", "mle"],2)}$ with confidence interval (CI) (\Sexpr{myround(ci_c["R0_be", "ci1"],2)},\Sexpr{myround(ci_c["R0_be", "ci2"],2)}) and $\Rzero^{\after}=\Sexpr{myround(ci_c["R0_af","mle"],2)}$  with CI (\Sexpr{myround(ci_c["R0_af", "ci1"],2)},\Sexpr{myround(ci_c["R0_af","ci2"],2)}), where the estimates and their associated 95\% CIs obtained by profile likelihood (Supplementary Sec.~\suppSecInference).
This implies that the Chinese government non-pharmaceutical interventions instituted on and around January 23 reduced $\Rzero$ by a factor of $\Sexpr{myround(ci_c["R0_be","mle"]/ci_c["R0_af","mle"],1)}$.
By contrast, the estimates of \cite{li20} are $\widetilde\Rzero^{\before}=2.38$  with CI $(2.03,2.77)$ and $\widetilde\Rzero^{\after}=0.98$  with CI $(0.83,1.16)$, implying reduction by a factor of $\Sexpr{myround(2.38/0.98,1)}$.
For comparison, interventions implemented across a panel of 41 countries (34 European) were estimated to reduce $\Rzero$ by a factor of $\Sexpr{myround(1/(1-0.77),1)}$  with CI $(\Sexpr{myround(1/(1-0.65),1)},\Sexpr{myround(1/(1-0.85),1)})$ \citep{brauner21}.
Our estimate for $\Rzero$ before lockdown is toward the high end of comparable estimates, reviewed by \citep{park20-jcm}.
An alternative metaopopulation analysis of the pre-lockdown China data, with a deterministic transmission model, obtained an $\Rzero$ estimate of 3.11  with CI  (2.39,4.13) \citep{read21}.

@@

\noindent {\it Model}.
Our model is a spatiotemporal partially observed Markov process (SpatPOMP) \citep{ionides21}.
A partially observed Markov process (POMP) \citep{breto09,king16} is a stochastic dynamic model comprised of: (i) a latent process having the Markov property that the future is conditionally independent of the past given the present; (ii) a measurement process, describing how the data are modeled as noisy and incomplete observations of the latent process.
The SpatPOMP framework adds an extra assumption that the latent process is comprised of a collection of units, each of which has its own latent process and observation process.
The latent processes for each unit can be interdependent, but the observations for a given unit are required to depend only on the latent process for that unit.
In our context, each city is a unit.
General notation for SpatPOMP models is set up in Supplementary Sec.~\suppSecOurModel, and this notation is subsequently used to define mathematically the specific SpatPOMP models studied in this paper (\LiMobility, \LiParams,\RevisedModelUnconstrained and \RevisedModelConstrained).
All the models under consideration have an SEAIR structure, as described in Figure~\ref{fig:flow_diagram}.

@@

\section*{Data availability}
The data are available as part of the \code{metapoppkg} R package.

\section*{Code availability}
The code generating the manuscript is archived on Zenodo
\eic{ZENODO MS LINK HERE} and a development version is at \url{https://github.com:jifanli/metapop_article}.
The software environment developed for this analysis is also available as both an archived version (\eic{ZENODO METAPOPPKG LINK HERE}) and a development version (\url{https://github.com:jifanli/metapoppkg}).
These programs depend on other open-source software archived at \url{https://cran.r-project.org/}.

\section*{Acknowledgements}
This work was supported by National Science Foundation grants DMS-1761603 and NSF DMS-1761612.

\section*{Author contributions}
J.L., E.L.I. and N.N. designed the study and carried out the numerical analysis. J.L., E.L.I, A.A.K. and N.N. developed the software environment supporting the data analysis. All authors discussed and interpreted the results. E.L.I. drafted the manuscript, and all authors edited the manuscript.

>>>>

\newcommand\arxiv[2]{#1}  %% for arxiv version
%% \newcommand\arxiv[2]{#2}  %% suppressing arxiv formatting

<<setup,echo=F,message=F,cache=F>>=
# computational intensity, i, also known as run level
i <- 3
library(metapoppkg)
Ncities <- ncol(metapoppkg::incidence) - 1 # first column is date
U <- switch(i,20,Ncities,Ncities)
library(foreach)
library(doParallel)
library(doRNG)
cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
if(is.na(cores)) cores <- detectCores()
registerDoParallel(cores)

@@

\begin{abstract}
Mathematical models in ecology and epidemiology must be consistent with observed data in order to generate reliable knowledge and sound policy. Metapopulation systems, which consist of a collection of sub-populations at various locations, pose technical challenges in statistical inference due to nonlinear, stochastic interactions. Difficulties encountered in these methodological issues can obstruct the core scientific questions concerning the link between the mathematical models and the data. Progress in statistically efficient simulation-based inference for partially observed stochastic dynamic systems has enabled the development of statistically rigorous approaches to the analysis of nonlinear but low-dimensional systems. Recently, an algorithm has been developed which enables comparable inference for higher-dimensional models arising in metapopulation systems. The COVID-19 pandemic provides a situation where mathematical models and their policy implications were widely visible, and we revisit an influential metapopulation model used to inform basic epidemiological understanding early in the pandemic. Our methods support self-critical data analysis, enabling us to identify and address weaknesses, and leading to a new model with substantially improved statistical fit and parameter identifiability. Our results suggest that the lockdown initiated on January 23, 2020 in China was more effective than previously thought. We proceed to recommend statistical analysis standards for future metapopulation system modeling.
\end{abstract}

@@

Biological populations may be structured into a collection of densely-populated communities separated by sparsely populated regions.
The communities, which may be cities in a human context, comprise a metapopulation.
Motivation for metapopulation modeling arises when some essential feature of the population dynamics cannot be understood from looking at a single location.
Dynamics of persistence through local extinctions and reintroductions have been extensively studied in ecology \cite{hanski98,mackenzie09}.
In epidemiology, metapopulation dynamics can be a barrier to the regional elimination and eventual eradication of a pathogen, and may determine the successful invasion of a new pathogen or a new strain of an existing pathogen \cite{metcalf21}.
In other situations, spatiotemporal dynamics may be an unavoidable component of the system under study without being the focus of the investigation \cite{zhang22,wheeler23}.

A recent growth in the study of metapopulation dynamics has been driven partly by the COVID-19 pandemic \cite{li20,wu20,wang22,prieto22,cascante-vega22,pizzuti20,alleman21,yang21,engebretsen23} and in part by methodological advances facilitating the fitting of metapopulation models to spatiotemporal data.
Until the start of this millenium, developing dynamic models with both statistical and scientific justification was a longstanding open problem for even a single community \cite{bjornstad01}.
Over the past two decades, new algorithms \cite{ionides06,toni09,andrieu10,ionides15} and software \cite{king16,kristensen16,devalpine17} together with ever-increasing computational resources, have enabled routine inference for low-dimensional nonlinear partially observed stochastic dynamic systems.
However, fundamental algorithmic scalability issues known as the ``curse of dimensionality'' lead to difficulties with the high-dimensional systems arising in metapopulation inference.
These issues are clearest for Monte Carlo techniques based on importance sampling \cite{bengtsson08} but are also evident in the need for variational approximations for large Monte Carlo Markov Chain (MCMC) calculations \cite{blei17}. 
The quest for statistical methods suitable for metapopulation models of scientific interest has motivated the development of various methodological approaches.
Nevertheless, limitations have persisted in the capability to carry out flexible model-based inference and rigorous model criticism.
Thus, data analysis for metapopulation models has lagged behind low-dimensional time series analysis of biological systems.
Recent developments enable this gap to be closed, as we demonstrate via a reanalysis of COVID-19, viewed from the context of the ability to drawm evidence-based scientific conclusions about the dynamics of the emerging pandemic in January and February 2020.

@@

A natural place to look for statistical methodology applicable to metapopulation models is among the techniques developed for inferring population dynamics at a single location, reviewed by \cite{funk20,auger-methe21}.
Commonly implemented approaches for POMP models can be categorized as  (i) MCMC; (ii) matching summary statistics between data and simulations; (iii)  linearization;   (iv) particle filters (i.e., sequential Monte Carlo).

In principle, MCMC techniques enable Bayesian inference or maximum likelihood via expectation-maximization algorithms \cite{cappe05}.
In practice, successful MCMC for metapopulation models requires careful model-specific algorithm development \cite{whitehouse23}.

Matching summary statistics of simulations to the corresponding data statistic is, in principle, a readily applicable inference approach for a wide class of models including metapopulation models.
In the context of Bayesian inference this is called Approximate Bayesian Computing \cite{conlan12}.
However, informative, low-dimensional summary statistics can be hard to construct even for low-dimensional nonlinear systems.
This can make summary statistic methods statistically inefficient \cite{fasiolo16,shrestha11}.

@@

For low-dimensional systems, particle filter methods are applicable to a flexible class of models for low-dimensional nonlinear systems \cite{breto09,king16}.
They have the attractive feature of statistically efficient use of data, i.e., accessing the likelihood function for Bayesian or likelihood-based inference, with approximation resulting only from finite Monte Carlo effort.
For high-dimensional systems, scalability considerations demand further approximations since particle filters suffer acutely from the ``curse of dimensionality'' \cite{bengtsson08}.
In contrast, block particle filter (BPF) methods achieve scalability by updating particles through localized resampling, rather than employing the linear update of EnKF \cite{rebeschini15}.
It is an empirical question which of these approximations is more successful on metapopulation models,  with prior evidence favoring BPF \cite{ionides21}.
In this paper, we not only aim to compare these methods in the context of metapopulation modeling but importantly also we reveal issues of model assessment when fitting complicated models to large datasets.
We introduce a model diagnosis procedure to mitigate the risk of selecting an inappropriate model, an essential step in informing public policy decisions.

@@

As an example, we reconsider the influential analysis of COVID-19 from early in the pandemic by Li et al. \cite{li20}.
This analysis provided estimates of transmission parameters and the effect of the lockdown in China using the limited data available at the time.
Other teams have fitted models to address similar questions \cite{kraemer20,yang21,brett23} but the study by \cite{li20} is distinctive for fitting a stochastic mechanistic metapopulation model to extensive spatiotemporal data.
The results were published in May, 2020, based on reported cases from January 10 to February 8 of that year.
The state-of-the-art spatiotemporal analysis was possible on an urgent timescale because the team of researchers had developed their methodology in a sequence of previous situations \cite{shaman12,yang14,yang15,pei18}.
The paper is written with attention to reproducibility, and the main results are strengthened by various supporting analyses in an extensive supplement.
While examining the points mentioned above, we have identified various limitations that could have been mitigated by adhering to the aforementioned recommendations.
Our goal is not to criticize any specific paper, but rather to build on the timely analysis of \cite{li20} to demonstrate how recently developed techniques provide possibilities to carry out improved data analysis in future.

@@

\end{tikzpicture}
}
\end{center}
\vspace{-5mm}
\caption{A flow diagram for the SEAIR metapopulation model.
Each individual in city $u$ is a member of exactly one of the square blue compartments.
Individuals entering the reportable infectious compartment, $\mathrm{I}_u$ for city $u$, are simultaneously included in the delayed reporting process compartment, $\mathrm{C^a_u}$.
Upon arrival at the final reporting compartment, $\mathrm{C}_u$, the individual is included in the case report for city $u$.
Individuals in $\mathrm{A}_u$ are not reportable and transmit at a reduced rate.
Movement of individuals between cities occurs by transport to and from a transport compartment, $\mathrm{T}$.
The number of individuals moving between each pair of cities is based on 2018 data from Tencent.
Movement is modeled only for susceptible, exposed, and undetected infections.}\label{fig:flow_diagram}
\end{figure}

For our metapopulation system, the sub-populations are 373 provincial cities in China (meaning cities with administrative responsibility for an entire region) and the data are daily reported COVID-19 cases. 
COVID-19 dynamics are represented by an Susceptible-Exposed-Asymptomatic-Infectious-Recovered (SEAIR) epidemic model. 
Questions of urgent interest early in the pandemic include the relative transmissibility of reported to unreported cases, the fraction of unreported cases, and the effect on transmission of movement restrictions imposed on and around January 23 \cite{li20}.
The model structure is illustrated by the flow diagram in Fig.~\ref{fig:flow_diagram}.
We consider different model implementations within this structure, with full model specifications provided by the equations and parameter values in Supplementary Sec.~\suppSecOurModel.

Our starting point is model {\LiMobility} which is based on the model of \cite{li20} and is fully described in Supplementary Sec.~\suppSecLiModel.
We consider the full dataset, from January 10 to February 8, with transmission parameters re-estimated following the lockdown on January 23; these correspond to the periods~1 (January 10 to January 22) and 3 (January 24 to February 8) of \cite{li20}.
Some minor differences between {\LiMobility} and \citep{li20} were introduced to enable us to place their model within the general metapopulation framework of spatiotemporal partially observed  continuous-time Markov processs described by \citep{ionides21}.
Despite these modifications, simulations from {\LiMobility}, using the parameters of \citep{li20}, closely match simulations from the code provided by \cite{li20} (Supplementary Sec.~\suppSecLiModel).
However, inspection of the mobility data reveals that some small cities have no recorded incoming travelers, and therefore no possibility of a SARS-CoV2 introduction within {\LiMobility} (or the model of \cite{li20}) (Supplementary Sec.~\suppSecMobility).
This minor limitation formally results in a likelihood of zero for {\LiMobility} (i.e., it is impossible for the simulation model to reproduce the observed spatiotemporal dataset), and hence a log-likelihood of $-\infty$.

@@

\begin{table}[ht]
\small
\begin{tabular}{c|rr|l}
Model & loglik & df & description
\\
\hline
{\LiMobility} &
  $-\infty$ & 10 &
  SEAIR model using the parameter values and mobility data of \cite{li20}
\\
{\LiParams} &
  \Sexpr{metapoppkg::myround(mean(M2_loglik['est']),1)}\stdError{\Sexpr{metapoppkg::myround(M2_loglik['se'],1)}}& 10 &
  Adjusted mobility and measurement in {\LiMobility}
\\
{\BenchmarkIID} &
  \Sexpr{metapoppkg::myround(-iid_mle3$value,1)}  & \Sexpr{U+1} &
  Independent identically distributed negative binomial
\\
{\BenchmarkAR} &
 \Sexpr{metapoppkg::myround(-ar_mle3$value,1)}  & \Sexpr{U+2} &
  Autoregressive negative binomial
\\
{\RevisedModelUnconstrained} &
  \Sexpr{metapoppkg::myround(lik3["logLik"],1)}\stdError{\Sexpr{metapoppkg::myround(lik3["logLik_se"],1)}}&
  12 &
  Adding overdispersed dynamics to {\LiParams} and refitting
\\
{\RevisedModelConstrained} &
  \Sexpr{metapoppkg::myround(lik6["logLik"],1)}\stdError{\Sexpr{metapoppkg::myround(lik6["logLik_se"],1)}}&
  10 &
  Latent and infectious durations unchanged by lockdown in {\RevisedModelUnconstrained}.
\\

@@

We addressed the problematic mobility data in {\LiMobility} by adding some additional transportation based on a gravity movement model, as described in Supplementary Sec.~\suppSecMobility, giving rise to model~{\LiParams}.
We implemented an additional adjustment between models {\LiMobility} and {\LiParams} to align the measurement model with the ensemble Kalman filter (EnKF) inference method presented by \cite{li20}.
That EnKF implementation involved specifying a quantity called the observation error variance, defined as a function of the observed cases, to quantify the uncertainty in the measurements.
Within the POMP specification, the measurement variance can depend on the latent state but not directly on the observed data.
To interpret the choice of EnKF observation variance within the POMP framework, we specified the measurement model for~{\LiParams} to have equivalent scaling to the choice of \citep{li20}, but with dependence on the reported cases replaced by dependence on the modeled, but unobserved, exact case count.

Based on a comparison of various nonlinear spatiotemporal filters (Supplemtary Fig.~\suppFigFilterComparison) we evaluated the log-likelihood for {\LiParams} using a block particle filter (Table~\ref{tab:loglik}).
To account for model overfitting, the number of estimated parameters can be subtracted from the log-likelihood to obtain a comparison equivalent to Akaike's Information Criterion (AIC) \citep{burnham02}.
When the difference in log-likelihood is large compared to the difference in degrees of freedom, the ordering of statistical goodness-of-fit is clear without presenting formal statistical hypothesis tests.

To find out whether this log-likelihood value suggests that the model is satisfactory, we compare it with two simple statistical models:
{\BenchmarkIID} simply models the daily case report for each city as an independent identically distributed (IID) negative binomial random variable; 
{\BenchmarkAR} adds an autoregressive component to {\BenchmarkIID} (see Supplemetary Sec.~\suppSecBenchmark).
We see from Table~\ref{tab:loglik} that both {\BenchmarkIID} and {\BenchmarkAR} outperform {\LiParams} by many units of log-likelihood.
Likelihood can properly be compared between different models for the same data, with statistical uncertainty in log-likelihood differences arising on the unit scale \citep{pawitan01}.
When the fit of a mechanistic model is inferior to a simple statistical model, we learn that the mechanistic model has room for improvement as a description of the data, but we do not immediately learn what the deficiency is.
The development of methods for formal statistical fitting of mechanistic models has led to increased understanding of the importance of appropriate modeling of over-dispersed variation in the stochastic dynamics \cite{he10,stocks20,whitehouse23}.
We therefore hypothesized that the fit of {\LiParams} could be improved by permitting additional dynamic noise.

@@

All parameter estimates are reported in Supplementary Table~\suppTableResults.
Here, we discuss the estimated basic reproductive number (i.e., the expected number of secondary infections from one index case in a fully susceptible population), denoted by $\Rzero^{\before}$ and $\Rzero^{\after}$ before and after the January 23rd lockdown.
Our estimates for model {\RevisedModelConstrained}, are $\Rzero^{\before}=\Sexpr{metapoppkg::myround(ci_c["R0_be", "mle"],2)}$ with confidence interval (CI) (\Sexpr{metapoppkg::myround(ci_c["R0_be", "ci1"],2)},\Sexpr{metapoppkg::myround(ci_c["R0_be", "ci2"],2)}) and $\Rzero^{\after}=\Sexpr{metapoppkg::myround(ci_c["R0_af","mle"],2)}$  with CI (\Sexpr{metapoppkg::myround(ci_c["R0_af", "ci1"],2)},\Sexpr{metapoppkg::myround(ci_c["R0_af","ci2"],2)}), where the estimates and their associated 95\% CIs obtained by profile likelihood (Supplementary Sec.~\suppSecInference).
This implies that the Chinese government non-pharmaceutical interventions instituted on and around January 23 reduced $\Rzero$ by a factor of $\Sexpr{metapoppkg::myround(ci_c["R0_be","mle"]/ci_c["R0_af","mle"],1)}$.
By contrast, the estimates of \cite{li20} are $\widetilde\Rzero^{\before}=2.38$  with CI $(2.03,2.77)$ and $\widetilde\Rzero^{\after}=0.98$  with CI $(0.83,1.16)$, implying reduction by a factor of $\Sexpr{metapoppkg::myround(2.38/0.98,1)}$.
For comparison, interventions implemented across a panel of 41 countries (34 European) were estimated to reduce $\Rzero$ by a factor of $\Sexpr{metapoppkg::myround(1/(1-0.77),1)}$  with CI $(\Sexpr{metapoppkg::myround(1/(1-0.65),1)},\Sexpr{metapoppkg::myround(1/(1-0.85),1)})$ \citep{brauner21}.
Our estimate for $\Rzero$ before lockdown is toward the high end of comparable estimates, reviewed by \citep{park20-jcm}.
An alternative metaopopulation analysis of the pre-lockdown China data, with a deterministic transmission model, obtained an $\Rzero$ estimate of 3.11  with CI  (2.39,4.13) \citep{read21}.

@@

\noindent {\it Model}.
Our model is a spatiotemporal partially observed Markov process (SpatPOMP) \citep{ionides21}.
A partially observed Markov process (POMP) \citep{breto09,king16} is a stochastic dynamic model comprised of: (i) a latent process having the Markov property that the future is conditionally independent of the past given the present; (ii) a measurement process, describing how the data are modeled as noisy and incomplete observations of the latent process.
The SpatPOMP framework adds an extra assumption that the latent process is comprised of a collection of units, each of which has its own latent process and observation process.
The latent processes for each unit can be interdependent, but the observations for a given unit are required to depend only on the latent process for that unit.
In our context, each city is a unit.
General notation for SpatPOMP models is set up in Supplementary Sec.~\suppSecOurModel, and this notation is subsequently used to define mathematically the specific SpatPOMP models studied in this paper ({\LiMobility}, {\LiParams},{\RevisedModelUnconstrained} and {\RevisedModelConstrained}).
All the models under consideration have an SEAIR structure, as described in Figure~\ref{fig:flow_diagram}.

@@

\arxiv{}{
\section*{Data availability}
The data are available as part of the \code{metapoppkg} R package.
}

\arxiv{}{
\section*{Code availability}
The code generating the manuscript is archived on Zenodo
\eic{ZENODO MS LINK HERE} and a development version is at \url{https://github.com:jifanli/metapop_article}.
The software environment developed for this analysis is also available as both an archived version (\eic{ZENODO METAPOPPKG LINK HERE}) and a development version (\url{https://github.com:jifanli/metapoppkg}).
These programs depend on other open-source software archived at \url{https://cran.r-project.org/}.
}

\section*{Acknowledgements}
This work was supported by National Science Foundation grants DMS-1761603 and DMS-1761612.
High performance research computing resources were provided by Texas A\&M University and University of Michigan.

\arxiv{}{
\section*{Author contributions}
J.L., E.L.I. and N.N. designed the study and carried out the numerical analysis. J.L., E.L.I, A.A.K. and N.N. developed the software environment supporting the data analysis. All authors discussed and interpreted the results. E.L.I. drafted the manuscript, and all authors edited the manuscript.
}
